Can We Formalise Type Theory Intrinsically without Any Compromise?

### 2025-12-03 tex/CPP26/main.tex
Remove leftover comment

% FNF (Fri 5 Sep)

@@

Early approaches to formalising type theory, e.g.\ McKinna and Pollack~\cite{McKinna1999}, dealt with untyped terms that were later refined by a typing relation, setoid equality, or quotients, and thus had to prove a lot of congruence lemmas by hand~\cite{Danielsson2006,Chapman2009,Brunerie2019,Boer2020}.
%
A breakthrough was achieved by Altenkirch and Kaposi~\cite{Altenkirch2016a}, who showed that quotient inductive-inductive types (QIITs)~\cite{Altenkirch2018} can be employed to significantly simplify the internal representation of well typed terms (or, more precisely, derivations), since equality constructors can be used to represent equations such as $\beta$- and $\eta$-equality.
%
They took Dybjer's notion of a model of type theory in the form of a category with families~\cite{Dybjer1996}, and translated it into a QIIT definition.
%
In effect, this gives rise to the \emph{initial} category with families, with the elimination principles of the QIIT giving a unique morphism of categories with families to any other model.
%
This thus gives a both principled and practical way to formalise the syntax and equational theory of type theory in type theory; the feasibility of the approach was demonstrated by e.g.\ formalising normalisation by evaluation using this representation~\cite{Altenkirch2017}.

@@

%
The framework of categories with families is only one of several (more or less) equivalent notions of models of type theory~\cite{Hofmann1997}, and we were wondering if any of the other notions might offer any advantages.
%
Bense~\cite{Bense2024} suggested that Awodey's notion of natural model~\cite{Awodey2018} might be a good candidate.
%
Indeed, in a natural model, the indexing of terms over their types $\mathsf{Tm}_{\Gamma} : \mathsf{Ty}(\Gamma) \to \mathsf{Set}$ (as in a category with families) is replaced by a `fibred' perspective where each term instead \emph{has} a type, as picked out by a function $\mathsf{tyOf} : \mathsf{Tm}(\Gamma) \to \mathsf{Ty}(\Gamma)$.
%
Terms and types are still indexed by contexts $\Gamma$, but since most `type mismatches' arise from equations between types, not equations between contexts (indeed many formulations of type theory does not even have a notion of context equality), this should mean that many uses of transports can be avoided.

We test this hypothesis by formalising type theory in a form inspired by natural models.
%
\CA is particularly a good fit for such a project, because not only does it support QIITs, it also supports inductive-recursive types~\cite{Dybjer1999}, which are needed to simultaneously define the recursive $\mathsf{tyOf}$ function together with the inductively defined types $\mathsf{Tm}(\Gamma)$ and $\mathsf{Ty}(\Gamma)$.
%
Indeed, it could be the lack of support for inductive-recursive definitions in many proof assistants which has held back formalisation attempts based on natural models so far.

While we manage to avoid transports occurring in its own syntax, the experiment is not an outright success.
%
Indeed, we found that when developing more sophisticated metatheory, such as when defining a logical predicate model, the use of transports along equations often reappeared.
%
Strictification~\cite{Donko2022,Kaposi2025}, a recent technique for strictifying equations, cannot be applied directly because of the lack of strict proposition in the metatheory.
%
Furthermore, we found that the use of natural models is less well supported in the Cubical Agda of today, compared to approaches based purely on QIITs.
%
This is because we are more reliant on the computational behaviour of the recursively defined $\mathsf{tyOf}$ function, and this behaviour is only available in `later' clauses, which leads to the need for hacks and tricks to work around this limitation.
%
We discuss proof assistant features and their helpfulness further towards the end of the paper, after presenting our formalisation.

\paragraph{Contributions} We make the following contributions:
\begin{itemize}
\item We present an intrinsically well typed representation of the syntax of type theory, inspired by Awodey's natural models (\cref{sec:tt}).
\item We derive elimination and recursion principles for the syntax (\cref{sec:tt:elim}), and show how the standard model and the term model are constructed (\cref{sec:standard-model}).
\item We discuss strictification constructions on models, and show that they also apply to our notion of natural models (\cref{sec:strictify}).
\item We develop normalisation by evaluation for substitution calculus~(\cref{sec:nbe}) as a proof of concept: our development is carried out in \CA, which has a computational implementation of QIITs and principles such as function extensionality, so the resulting normaliser computes, and can potentially be extracted as a verified program.
\item We discuss pros and cons of our approach compared to other approaches, and which features of a proof assistant and its metatheory would make future formalisation more feasible (\cref{sec:discussion}).
\end{itemize}

%\LT{the idea of using natural model appears at least in 2024 \cite{Bense2024}, and it is a natural idea to formalise type theory in this way.}
\section{Setting and metatheory}
% FNF (Sun 7 Sep)

%\LT{Should we discharge this assumption in our formalisation, eg set-truncate where needed instead? (R1, R2)}
Our formalisation is carried out in \CA without the use of the \ensuremath{\Conid{Glue}} type and, in particular, the univalence principle, checked using the option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} available in the forthcoming \Agda 2.9.0.
%
We explicitly set-truncate the types we define.
Therefore, the term \emph{set} is used interchangeably with type.
%
For the sake of simplicity we occasionally postulate uniqueness of identity proofs (UIP) locally, which is of course inconsistent with univalence but in principle compatible with cubical type theory. 
%
We believe that the cubical type theory XTT~\cite{Sterling2022}, which enjoys definitional UIP without univalence, justifies this local assumption.
%

@@

In this section, we will explain why Altenkirch and Kaposi's representation~\cite{Altenkirch2016a} is hard to use in practice and rejected by \CA arising from transports in its definition.
Then, we show how their representation can be transformed to a representation based on Awodey's natural models.
This representation is accepted by \CA, since it is free from transports.

@@

In other words, transport hell is not only an obstacle for reasoning, but also breaks strict positivity in \CA when arising in inductive definitions themselves.
The situation becomes worse once additional type formers are introduced --- such as $\Pi$-types and the type \ensuremath{\Conid{El}} of elements~\cite{Altenkirch2016a} --- since each brings further instances of this problem.
%
Of course, one could bypass the strict positivity check, but doing so would undermine the general trustworthiness of formalisation.
Another possibility is to fix the syntactic restriction for HIITs, but it is unclear what conditions should be.
Therefore, we seek for an equivalent definition without transports first.

Another source of transports arises from equations over equations, but this can be avoided by using dependent paths.
For example, the fact that the identity term substitution really acts as an identity is introduced as an equality constructor \ensuremath{[\mathsf{idS}]_{\text{t}}}, defined over the highlighted equality constructor \ensuremath{[\mathsf{idS}]_{\text{T}}\;\mathbin{:}\;\Conid{A}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}} for the identity type substitution:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}[\mathsf{idS}]_{\text{t}}\;\mathbin{:}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}\;(\highlight{addition}{[\mathsf{idS}]_{\text{T}}\;\Varid{i}}))\;\Varid{t}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{t}}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Although equations over equations are in principle more manageable, it quickly leads us to \emph{equations over equations over yet more equations} in their elimination rules.
Hence, it is still preferable to avoid them if possible.

@@

To avoid transports in the definition itself, we note that the index \ensuremath{\Conid{A}} of \ensuremath{\Conid{Tm}\;\Conid{Γ}\;\Conid{A}} often needs an explicit proof for the typing constraint --- for example, that the term \ensuremath{\Varid{t}} in the substitution \ensuremath{(\Varid{σ}\;\Varid{,}\;\Varid{t})} has type \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} --- if this does not happen to hold strictly (i.e., up to judgemental equality), so enforcing this constraint in the index of \ensuremath{\Conid{Tm}} just shoots ourselves in the foot.
Hence, we apply the `Ford transformation'~\cite{McBride1999} (`You can have any index you want, as long as it is equal to the specified one') to move the constraint on its index to its argument as an equality proof (highlighted below):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\char95 ,\char95 ∶[\char95 ]}\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;\Conid{B})\;\highlight{addition}{(\Varid{pt}\;\mathbin{:}\;\Conid{B}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The constructor \ensuremath{\Varid{,∘}} which had a transport in its type becomes
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{,∘}\;{}\<[5]%
\>[5]{}\mathbin{:}\;\Varid{...}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;{}\<[E]%
\\
\>[5]{}\hsindent{6}{}\<[11]%
\>[11]{}\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\highlight{addition}{\Varid{cong}\;\Varid{\char95 [}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\Varid{pt}\;\Varid{∙}\;[\circ]_{\text{T}}}\;\mskip1.5mu]){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\Varid{\char95 ∙\char95 }} is the transitivity of equality.
Although transport is not needed this time, the use of \ensuremath{\Varid{cong}} and \ensuremath{\Varid{\char95 ∙\char95 }}
still prevent the definition from being seen as strictly positive.
Similar to the Ford transformation, this problem can be overcome by asking for another equality proof, highlighted below, as an argument:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{,∘}\;\mathbin{:}\;\Varid{...}\;\highlight{addition}{(\Varid{qt}\;\mathbin{:}\;\Conid{B}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Varid{→}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ})\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\highlight{addition}{\Varid{qt}}\;\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As \ensuremath{\Conid{Sub}} is a set, the additional argument is essentially unique, so this updated constructor does not require any information but only defers the proof obligation.
%This redundant argument can be removed later when defining its eliminator (\Cref{sec:tt:elim}).

@@

\subsection{Representing substitution calculus using QIIRT}
Building on the changes described in \Cref{sec:tt:terms-without-indices}, we now spell out our version of substitution calculus. The following types are defined simultaneously with a recursive function (changes compared to the QIIT version highlighted):

@@

We write \ensuremath{\Varid{wk}} for the `weakening' substitution \ensuremath{\Varid{wk}\;\mathrel{=}\;\Varid{π₁}\;\Varid{idS}\;\mathbin{:}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})\;\Conid{Γ}}, and \ensuremath{\Varid{vz}} for the `top variable' \ensuremath{\Varid{vz}\;\mathrel{=}\;\Varid{π₂}\;\Varid{idS}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})}.
This works well except that we have to interleave the function clauses of \ensuremath{\Varid{tyOf}} with constructors.

@@

Apart from the extra clauses of \ensuremath{\Varid{tyOf}}, the main change happens in the constructor \ensuremath{\Varid{app}}.
The constraint that \ensuremath{\Varid{t}} is of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} is enforced there, but every other constructor remains as almost the same as their QIIT definition.
We have formulated application and abstraction as an isomorphism between terms of type \ensuremath{\Conid{B}} in context \ensuremath{\Conid{Γ}\;\Varid{,}\;\Conid{A}} and terms of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} in context \ensuremath{\Conid{Γ}}, but we can also derive ordinary application:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\char95 ∶[\char95 ]\$\$\char95 ∶[\char95 ]}\;{}\<[15]%
\>[15]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\Varid{→}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{Π}\;\Conid{A}\;\Conid{B}\;{}\<[E]%
\\
\>[15]{}\Varid{→}\;(\Varid{s}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\Varid{→}\;\Varid{tyOf}\;\Varid{s}\;\mathop\equiv\;\Conid{A}\;{}\<[E]%
\\
\>[15]{}\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
\\
\>[B]{}\Varid{t}\;\Varid{∶[}\;\Varid{p}\;\Varid{]\$\$}\;\Varid{s}\;\Varid{∶[}\;\Varid{q}\;\mskip1.5mu]\;\mathrel{=}\;\Varid{app}\;\Varid{t}\;\anonymous \;\Varid{p}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{s}\;\Varid{∶[}\;\Varid{q}\;\Varid{∙}\;[\mathsf{idS}]_{\text{T}}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{t}}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

>>>>

@@

Early approaches to formalising type theory, e.g.\ McKinna and Pollack~\cite{McKinna1999}, dealt with untyped terms that were later refined by a typing relation or setoid equality, and thus had to prove a lot of congruence lemmas by hand~\cite{Danielsson2006,Chapman2009}.
%
A breakthrough was achieved by Altenkirch and Kaposi~\cite{Altenkirch2016a}, who showed that quotient inductive-inductive types (QIITs)~\cite{Altenkirch2018} can be employed to significantly simplify the internal representation of well typed terms (or, more precisely, derivations), since equality constructors can be used to represent equations such as $\beta$- and $\eta$-equality.
%
They took Dybjer's notion of a model of type theory in the form of a category with families~\cite{Dybjer1996}, and translated it into a QIIT definition.
%
In effect, this gives rise to the \emph{initial} category with families, with the elimination principles of the QIIT giving a unique morphism of categories with families to any other model.
%
This thus gives a both principled and practical way to formalise the syntax and equational theory of type theory in type theory; the feasibility of the approach was demonstrated by e.g.\ formalising normalisation by evaluation using this representation~\cite{Altenkirch2017}.

@@

%
The framework of categories with families is only one of several (more or less) equivalent notions of models of type theory~\cite{Hofmann1997}, and we were wondering if any of the other notions might offer any advantages.
%
Bense~\cite{Bense2024} suggested that Awodey's notion of natural model~\cite{Awodey2018} might be a good candidate.
%
Indeed, in a natural model, the family of terms indexed by their types $\mathsf{Tm}_{\Gamma} : \mathsf{Ty}(\Gamma) \to \mathsf{Set}$ (as in a category with families) is replaced by a `fibred' perspective where each term instead \emph{has} a type, as picked out by a function $\mathsf{tyOf} : \mathsf{Tm}(\Gamma) \to \mathsf{Ty}(\Gamma)$.
%
Terms and types are still indexed by contexts $\Gamma$, but since most `type mismatches' arise from equations between types, not equations between contexts (indeed many formulations of type theory do not even have a notion of context equality), this should mean that many uses of transports can be avoided.

We test this hypothesis by formalising type theory in a form inspired by natural models.
%
\CA is a particularly good fit for such a project, because not only does it support QIITs, it also supports inductive-recursive types~\cite{Dybjer1999}, which are needed to simultaneously define the recursive $\mathsf{tyOf}$ function together with the inductively defined types $\mathsf{Tm}(\Gamma)$ and $\mathsf{Ty}(\Gamma)$.
%
Indeed, it could be the lack of support for inductive-recursive definitions in many proof assistants which has held back formalisation attempts based on natural models so far.

While we manage to avoid transports occurring in its own syntax, the experiment is not an outright success.
%
Indeed, we found that when developing more sophisticated metatheory, such as when defining a logical predicate model, the use of transports along equations often reappeared.
%
Strictification~\cite{Donko2022,Kaposi2025} --- a recent technique for making certain equations hold up to definitional equality --- partially helps, but does not completely eliminate the problem.
%
Furthermore, we found that the use of natural models is less well supported in the Cubical Agda of today, compared to approaches based purely on QIITs.
%
This is because we are more reliant on the computational behaviour of the recursively defined $\mathsf{tyOf}$ function, and this behaviour is only available in `later' clauses, which leads to the need for hacks and tricks to work around this limitation.
%
We discuss proof assistant features and their helpfulness further towards the end of the paper, after presenting our formalisation.

\paragraph{Contributions} We make the following contributions:
\begin{itemize}
\item We present an intrinsically well typed representation of the syntax of type theory, inspired by Awodey's natural models (\cref{sec:tt}).
\item We derive elimination and recursion principles for the syntax (\cref{sec:tt:elim}), and show how the standard model and the term model are constructed (\cref{sec:standard-model}).
\item We discuss strictification constructions on models, and show that they also apply to our notion of natural models (\cref{sec:strictify}).
\item We develop normalisation by evaluation for the substitution calculus~(\cref{sec:nbe}) as a proof of concept: our development is carried out in \CA, which has a computational implementation of QIITs and principles such as function extensionality, so the resulting normaliser computes, and can potentially be extracted as a verified program.
\item We discuss pros and cons of our approach compared to other approaches, and which features of a proof assistant and its metatheory would make future formalisation more feasible (\cref{sec:discussion}).
\end{itemize}

\section{Setting and metatheory}

%\LT{Should we discharge this assumption in our formalisation, eg set-truncate where needed instead? (R1, R2)}
Our formalisation is carried out in \CA without the use of the \ensuremath{\Conid{Glue}} type and, in particular, the univalence principle, checked using the option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} available in the forthcoming \Agda 2.9.0. It also typechecks with \Agda version 2.8.0.
%
We explicitly set-truncate the types we define.
Therefore, the term \emph{set} is used interchangeably with type.
%
For the sake of simplicity we occasionally postulate uniqueness of identity proofs (UIP) locally, which is of course inconsistent with univalence but in principle compatible with cubical type theory.
%
We believe that the cubical type theory XTT~\cite{Sterling2022}, which enjoys definitional UIP without univalence, justifies this local assumption.
%

@@

In this section, we explain why Altenkirch and Kaposi's representation~\cite{Altenkirch2016a} is hard to use in practice, and rejected by \CA due to transports in its definition.
Then, we show how their representation can be transformed to a representation based on Awodey's natural models.
This representation is accepted by \CA, since it is free from transports.

@@

In other words, transport hell is not only an obstacle for reasoning, but also breaks strict positivity in \CA when arising in inductive definitions themselves.
The situation becomes worse once additional type formers are introduced --- such as $\Pi$-types and the type \ensuremath{\Conid{El}} of elements~\cite{Altenkirch2016a} --- since each brings further instances of this problem.
%
Of course, one could bypass the strict positivity check, but doing so would undermine the general trustworthiness of formalisation.
Another possibility is to fix the syntactic restriction for HIITs, but it is unclear what the conditions should be.
Therefore, we seek an equivalent definition without transports first.

One option is to use \emph{dependent paths} to encode equations over equations.
For example, the fact that the identity term substitution really acts as an identity is introduced as an equality constructor \ensuremath{[\mathsf{idS}]_{\text{t}}}, defined over the highlighted equality constructor \ensuremath{[\mathsf{idS}]_{\text{T}}\;\mathbin{:}\;\Conid{A}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{T}}} for the identity type substitution:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}[\mathsf{idS}]_{\text{t}}\;\mathbin{:}\;\Conid{PathP}\;(\Varid{λ}\;\Varid{i}\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}\;(\highlight{addition}{[\mathsf{idS}]_{\text{T}}\;\Varid{i}}))\;\Varid{t}\;(\Varid{t}\;[\kern-1.5pt\;\Varid{idS}\;\kern-1.5pt]_{\text{t}}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
This is in principle managable --- indeed, Altenkirch, Kaposi and Xie~\cite{Altenkirch2026} successfully use dependent paths and some encoding tricks to represent the groupoid syntax of type theory in \CA --- but the \ensuremath{\Conid{PathP}} type is not part of a standard presentation of type theory, but instead rather \CA specific (of course, dependent paths can also be encoded using transports, but then inheriting the problems described above).
Dependent paths also quickly leads to equations over equations over yet more equations in their elimination rules.
Hence, it is still preferable to avoid them if possible.

@@

To avoid transports in the definition itself, we note that the index \ensuremath{\Conid{A}} of \ensuremath{\Conid{Tm}\;\Conid{Γ}\;\Conid{A}} often needs an explicit proof for the typing constraint --- for example, that the term \ensuremath{\Varid{t}} in the substitution \ensuremath{(\Varid{σ}\;\Varid{,}\;\Varid{t})} has type \ensuremath{\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}}} --- if this does not happen to hold strictly (i.e., up to judgemental equality), so enforcing this constraint in the index of \ensuremath{\Conid{Tm}} just shoots ourselves in the foot.
Hence, we apply the `Ford transformation'~\cite{McBride1999} (``You can have any index you want, as long as it is equal to the specified one'') to move the constraint on its index to its argument as an equality proof (highlighted below):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\char95 ,\char95 ∶[\char95 ]}\;\mathbin{:}\;(\Varid{σ}\;\mathbin{:}\;\Conid{Sub}\;\Conid{Γ}\;\Conid{Δ})\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ}\;\Conid{B})\;\highlight{addition}{(\Varid{pt}\;\mathbin{:}\;\Conid{B}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{→}\;\Conid{Sub}\;\Conid{Γ}\;(\Conid{Δ}\;\Varid{,}\;\Conid{A}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The constructor \ensuremath{\Varid{,∘}} which had a transport in its type becomes
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{,∘}\;{}\<[5]%
\>[5]{}\mathbin{:}\;\Varid{...}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;{}\<[E]%
\\
\>[5]{}\hsindent{6}{}\<[11]%
\>[11]{}\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\highlight{addition}{\Varid{cong}\;\Varid{\char95 [}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\Varid{pt}\;\Varid{∙}\;[\circ]_{\text{T}}}\;\mskip1.5mu]){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\Varid{\char95 ∙\char95 }} is the transitivity of equality.
Although transport is not needed this time, the use of \ensuremath{\Varid{cong}} and \ensuremath{\Varid{\char95 ∙\char95 }}
still prevent the definition from being seen as strictly positive.
Similar to the Ford transformation, this problem can be overcome by asking for another equality proof, highlighted below, as an argument:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{,∘}\;\mathbin{:}\;\Varid{...}\;\highlight{addition}{(\Varid{qt}\;\mathbin{:}\;\Conid{B}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{T}}\;\mathop\equiv\;\Conid{A}\;[\kern-1.5pt\;\Varid{σ}\;\Varid{∘}\;\Varid{τ}\;\kern-1.5pt]_{\text{T}})}\;{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Varid{→}\;(\Varid{σ}\;\Varid{,}\;\Varid{t}\;\Varid{∶[}\;\Varid{pt}\;\mskip1.5mu])\;\Varid{∘}\;\Varid{τ}\;\mathop\equiv\;(\Varid{σ}\;\Varid{∘}\;\Varid{τ})\;\Varid{,}\;\Varid{t}\;[\kern-1.5pt\;\Varid{τ}\;\kern-1.5pt]_{\text{t}}\;\Varid{∶[}\;\highlight{addition}{\Varid{qt}}\;\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
As \ensuremath{\Conid{Sub}} is a set, the additional argument is essentially unique, so this updated constructor does not require any additional information but only defers the proof obligation.
%This redundant argument can be removed later when defining its eliminator (\Cref{sec:tt:elim}).

@@

\subsection{Representing the substitution calculus using QIIRT}
Building on the changes described in \Cref{sec:tt:terms-without-indices}, we now spell out our version of the substitution calculus. The following types are defined simultaneously with a recursive function (changes compared to the QIIT version highlighted):

@@

We write \ensuremath{\Varid{wk}} for the `weakening' substitution \ensuremath{\Varid{wk}\;\mathrel{=}\;\Varid{π₁}\;\Varid{idS}\;\mathbin{:}\;\Conid{Sub}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})\;\Conid{Γ}}, and \ensuremath{\Varid{vz}} for the most recently bound `zeroth variable' \ensuremath{\Varid{vz}\;\mathrel{=}\;\Varid{π₂}\;\Varid{idS}\;\mathbin{:}\;\Conid{Tm}\;(\Conid{Γ}\;\Varid{,}\;\Conid{A})}.

The above definition works well, except that we have to interleave the function clauses of \ensuremath{\Varid{tyOf}} with constructors.

@@

Apart from the extra clauses of \ensuremath{\Varid{tyOf}}, the main change happens in the constructor \ensuremath{\Varid{app}}.
The constraint that \ensuremath{\Varid{t}} is of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} is enforced there, but every other constructor remains almost the same as their QIIT definition.
We have formulated application and abstraction as an isomorphism between terms of type \ensuremath{\Conid{B}} in context \ensuremath{\Conid{Γ}\;\Varid{,}\;\Conid{A}} and terms of type \ensuremath{\Conid{Π}\;\Conid{A}\;\Conid{B}} in context \ensuremath{\Conid{Γ}}, but we can also derive ordinary application:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{\char95 ∶[\char95 ]\$\$\char95 ∶[\char95 ]}\;{}\<[15]%
\>[15]{}\mathbin{:}\;(\Varid{t}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\Varid{→}\;\Varid{tyOf}\;\Varid{t}\;\mathop\equiv\;\Conid{Π}\;\Conid{A}\;\Conid{B}\;{}\<[E]%
\\
\>[15]{}\Varid{→}\;(\Varid{s}\;\mathbin{:}\;\Conid{Tm}\;\Conid{Γ})\;\Varid{→}\;\Varid{tyOf}\;\Varid{s}\;\mathop\equiv\;\Conid{A}\;{}\<[E]%
\\
\>[15]{}\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
\\
\>[B]{}\Varid{t}\;\Varid{∶[}\;\Varid{p}\;\Varid{]\$\$}\;\Varid{s}\;\Varid{∶[}\;\Varid{q}\;\mskip1.5mu]\;\mathrel{=}\;\Varid{app}\;\Varid{t}\;\anonymous \;\Varid{p}\;[\kern-1.5pt\;\Varid{idS}\;\Varid{,}\;\Varid{s}\;\Varid{∶[}\;\Varid{q}\;\Varid{∙}\;[\mathsf{idS}]_{\text{T}}\;\mskip1.5mu]\;\kern-1.5pt]_{\text{t}}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

As an example, we can write the identity function at type \ensuremath{\Conid{A}} as \ensuremath{\Varid{id}\;\Conid{A}\;\mathrel{=}\;\Varid{abs}\;\{\mskip1.5mu \Conid{A}\;\mathrel{=}\;\Conid{A}\mskip1.5mu\}\;\Varid{vz}}, and the identity function at type \ensuremath{\Conid{Π}\;\Conid{A}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{wk}\;\kern-1.5pt]_{\text{T}})} applied to the identity function at type \ensuremath{\Conid{A}} is represented by
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{idid}\;\mathbin{:}\;(\Conid{A}\;\mathbin{:}\;\Conid{Ty}\;\Conid{Γ})\;\Varid{→}\;\Conid{Tm}\;\Conid{Γ}{}\<[E]%
\\
\>[B]{}\Varid{idid}\;\Conid{A}\;\mathrel{=}\;\Varid{id}\;(\Conid{Π}\;\Conid{A}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{wk}\;\mskip1.5mu]))\;\Varid{∶[}\;\Varid{refl}\;\Varid{]\$\$}\;\Varid{id}\;\Conid{A}\;\Varid{∶[}\;\Varid{refl}\;\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
and indeed using the equality constructors, we have a proof of \ensuremath{\Varid{tyOf}\;(\Varid{idid}\;\Conid{A})\;\mathop\equiv\;\Conid{Π}\;\Conid{A}\;(\Conid{A}\;[\kern-1.5pt\;\Varid{wk}\;\kern-1.5pt]_{\text{T}})} as expected.
